Perihelion advance of Mercury

### 2020-09-25 perihelion_advance_of_mercury.Rmd
label each chunk and use cache

```{r}
# Simulate a trajectory starting at r=30, mass of sun=1/2 so Schwarzschild radius=1:
jj <- trajectory(r_start=30, m=1/2, h=3) # h=3 gives angular momentum
plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, xlab='', ylab='',main='Geodesics in the Schwarzschild geometry')
polargrid(r=10*(1:3))
event_horizon()  # show the event horizon, to scale
```

@@

```{r}
jj <- trajectory(30, 1/2, 3, GR=FALSE) # GR=FALSE means use Newtonian approximation
plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, xlab='', ylab='',main='Newtonian approximation')
polargrid(r=10*(1:3))
event_horizon()
```

@@

First we specify some orbital parameters for the orbit of Mercury, mostly taken from Wikipedia:
```{r}
G <-  6.674e-11     # big G in SI
M  <- 1.9885e30     # mass of Sun (kg)
M_merc <- 3.3011e23 # mass of Mercury (kg)
sol <- 299792458    # speed of light (m/s)
merc_perihelion <- 46001200e3 
merc_aphelion   <- 69816900e3
mu <- G*(M + M_merc)
```

Now we calculate some features of the orbit in geometrized units:
```{r} 
d <- G*M/sol^2  # mass of sun in meters
r_s <- 2*d      # Schwarzschild radius of Sun (=2M)
```

And some further features using classical geometry:
```{r}
a <- (merc_perihelion + merc_aphelion)/2
cee <- a-merc_perihelion
b <- sqrt(a^2-cee^2)
ecc <- sqrt(1-(b/a)^2)  # should match eccentricity given by Wikipedia
ell <- (a^2-cee^2)/a  # semi-latus rectum
```

@@

```{r}
vmin <- sqrt(mu/ell)*(1-ecc)
vmax <- sqrt(mu/ell)*(1+ecc)
```

@@

```{r}   
specific_angular_momentum_SI <- merc_perihelion * vmax
h <- specific_angular_momentum_SI/sol #geometrized units
```

With all these we can simulate the orbit:
```{r}
 out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e15,len=5e5))
 plot(out$x,out$y,type='l',asp=1,axes=FALSE,xlab='',ylab='',main='Orbit of Mercury, Sun not to scale')
 points(0,0,pch=21,bg='yellow',col='orange') # Schwarzschild radius invisible at this scale
 max(out$phi/(2*pi))
```

@@

```{r}
out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7))
```

@@

```{r}
jj1 <- c(NA,out$rdot)
jj2 <- c(out$rdot,NA)
change <- (jj1>0) & (jj2<0)

@@

```{r}
prec <- lm(phi~tee)$coefficients[2]
prec
```

And to convert to astronomical units we use
```{r}
conversion <-
  c(
      seconds_per_minute = 60,
      minutes_per_hour   = 60,
      hours_per_day      = 24,
      days_per_year      = 365.25,
      years_per_century  = 100,
      degrees_per_radian = 180/pi,
      minutes_per_degree = 60,
      seconds_per_minute = 60
  )
```

@@

```{r}
prec*prod(conversion)
```

@@

```{r}
out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7),GR=FALSE)
jj1 <- c(NA,out$rdot)
jj2 <- c(out$rdot,NA)
change <- (jj1>0) & (jj2<0)
tee    <- out$time[change]
phi <- out$phi[change]
phi <- (phi) %% (2*pi)
plot(tee,phi-pi,main='orbital angle at of Mercury at perihelion, classical case',xlab='time (seconds)',ylab='angle of perihelion')
```

@@

```{r}
precession <- 24*pi^3*a^2/(per^2*sol^2*(1-ecc^2))  # radians/orbit
precession
```

@@

```{r}
seconds_per_orbit <- prod(c(
    seconds_per_minute = 60,
    minutes_per_hour   = 60,
    hours_per_day      = 24,
    days_per_orbit     = 87.969
))

>>>>

```{r,label=geodesicsimulatorGR}
# Simulate a trajectory starting at r=30, mass of sun=1/2 so Schwarzschild radius=1:
jj <- trajectory(r_start=30, m=1/2, h=3) # h=3 gives angular momentum
plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, xlab='', ylab='',main='Geodesics in the Schwarzschild geometry')
polargrid(r=10*(1:3))
event_horizon()  # show the event horizon, to scale
```

@@

```{r,label=geodesicsimulatorclassical}
jj <- trajectory(30, 1/2, 3, GR=FALSE) # GR=FALSE means use Newtonian approximation
plot(jj$x,jj$y, asp=1, type='l', axes=FALSE, xlab='', ylab='',main='Newtonian approximation')
polargrid(r=10*(1:3))
event_horizon()
```

@@

First we specify some orbital parameters for the orbit of Mercury, mostly taken from Wikipedia:

```{r,label=mercuryorbitalparameters}
G <-  6.674e-11     # big G in SI
M  <- 1.9885e30     # mass of Sun (kg)
M_merc <- 3.3011e23 # mass of Mercury (kg)
sol <- 299792458    # speed of light (m/s)
merc_perihelion <- 46001200e3 
merc_aphelion   <- 69816900e3
mu <- G*(M + M_merc)
```

Now we calculate some features of the orbit in geometrized units:

```{r,label=geometrizedunits} 
d <- G*M/sol^2  # mass of sun in meters
r_s <- 2*d      # Schwarzschild radius of Sun (=2M)
```

And some further features using classical geometry:

```{r,label=furtherfeaturesclassical}
a <- (merc_perihelion + merc_aphelion)/2
cee <- a-merc_perihelion
b <- sqrt(a^2-cee^2)
ecc <- sqrt(1-(b/a)^2)  # should match eccentricity given by Wikipedia
ell <- (a^2-cee^2)/a  # semi-latus rectum
```

@@

```{r,label=maxandminorbitalperiods}
vmin <- sqrt(mu/ell)*(1-ecc)
vmax <- sqrt(mu/ell)*(1+ecc)
```

@@

```{r,label=orbitalangularmomentumSI}   
specific_angular_momentum_SI <- merc_perihelion * vmax
h <- specific_angular_momentum_SI/sol #geometrized units
```

With all these we can simulate the orbit:
```{r,label=orbitsimulatorincludingGR,cache=TRUE}
 out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e15,len=5e5))
 plot(out$x,out$y,type='l',asp=1,axes=FALSE,xlab='',ylab='',main='Orbit of Mercury, Sun not to scale')
 points(0,0,pch=21,bg='yellow',col='orange') # Schwarzschild radius invisible at this scale
 max(out$phi/(2*pi))
```

@@

```{r,label=orbitsimulatorGRlongtime,cache=TRUE}
out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7))
```

@@

```{r,label=orbitalangleatperihelion}
jj1 <- c(NA,out$rdot)
jj2 <- c(out$rdot,NA)
change <- (jj1>0) & (jj2<0)

@@

```{r,label=calculateprecession}
prec <- lm(phi~tee)$coefficients[2]
prec
```

And to convert to astronomical units we use

```{r,label=converttoastronomialunits}
conversion <-
  c(
      seconds_per_minute = 60,
      minutes_per_hour   = 60,
      hours_per_day      = 24,
      days_per_year      = 365.25,
      years_per_century  = 100,
      degrees_per_radian = 180/pi,
      minutes_per_degree = 60,
      seconds_per_minute = 60
  )
```

@@

```{r,label=resultofprecessionconverted}
prec*prod(conversion)
```

@@

```{r,precessionwithclassicaldynamics,cache=TRUE}
out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7),GR=FALSE)
jj1 <- c(NA,out$rdot)
jj2 <- c(out$rdot,NA)
change <- (jj1>0) & (jj2<0)
tee    <- out$time[change]
phi <- out$phi[change]
phi <- (phi) %% (2*pi)
plot(tee,phi-pi,main='orbital angle at of Mercury at perihelion, classical case',xlab='time (seconds)',ylab='angle of perihelion')
```

@@

```{r,label=theoreticalprecessioncalculator}
precession <- 24*pi^3*a^2/(per^2*sol^2*(1-ecc^2))  # radians/orbit
precession
```

@@

```{r,label=converttohumancentricunits}
seconds_per_orbit <- prod(c(
    seconds_per_minute = 60,
    minutes_per_hour   = 60,
    hours_per_day      = 24,
    days_per_orbit     = 87.969
))
